On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-03-29 manuscripts/PNAS/chazthm_pnas.tex
merge

% an important contribution of our work is to
Next, we identify the combinatorial criteria on support sets of generating codes allowing for stable sparse 
representations.  Let $[m]$ denote the set $\{1, \ldots, m\}$, its power set by $2^{[m]}$, and ${[m] \choose k}$ the set of subsets of $[m]$ of size $k$. Let $\text{\rm Span}\{\mathbf{v}_1, \ldots, \mathbf{v}_\ell\}$ be the $\mathbb{R}$-linear span of vectors $\mathbf{v}_1, \ldots, \mathbf{v}_\ell$. A vector $\mathbf{x}$ is said to be \emph{supported} on $S \subseteq [m]$, written supp$(\mathbf{x}) = S$, when $\mathbf{x} \in \text{\rm Span} \{\mathbf{e}_j\}_{j\in S}$, where $\mathbf{e}_j$ are the standard basis elements in $\mathbb R^m$.  We denote by $\mathbf{M}_j$ the $j$th column of a matrix $\mathbf{M}$ and similarly by $\mathbf{M}_S$ the sub-matrix formed by the columns of $\mathbf{M}$ indexed by $S$ ($\text{\rm Span}\{\mathbf{M}_\emptyset\} := \{\textbf{0}\}$).  We say a hypergraph $E \subseteq 2^{[m]}$ is of \textit{rank} $k$ when every $S \in E$ has cardinality $|S| \leq k$ (and \textit{$k$-uniform} when equality holds). We say $E$ is \emph{regular} when every element of $[m]$ is contained in exactly $\ell$ elements of $E$ for some $\ell > 0$ (for given $\ell$, we say $E$ is $\ell$-regular).

\begin{definition}
Given a hypergraph $E \subseteq 2^{[m]}$, the \textbf{star} $F(i)$ centered at $i$ is the set of $S \in E$ containing $i$. We say $E$ satisfies the \textbf{singleton intersection property} (SIP) when $\cap_{S \in F(i)} S = \{i\}$ for all $i \in [m]$. 
\end{definition}

@@

\begin{theorem}\label{DeterministicUniquenessTheorem}
% Fix positive integers $k, \ell, m, m', n, p$ satisfying $k < m$ and $p \leq m \leq m' \leq (\ell m - p) / (\ell - 1)$. 
Fix integers $n, \ell$ and $k < m \leq \hat m$. Suppose $\mathbf{A}~\in~\mathbb{R}^{n \times m}$ satisfies the spark condition \eqref{SparkCondition} and that $k$-sparse \mbox{$\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$} contain at least \mbox{$(k-1){\hat m \choose k}+1$} vectors supported in general linear position on each set in some $\ell$-regular $E \subseteq {[m] \choose k}$ satisfying the SIP.  There exists a constant $C_1 > 0$ for which the following holds for all $\varepsilon < L_2(\mathbf{A}) / C_1$.

@@

In other words, the smaller the difference $\hat m - m$, the more columns and coefficients of the original $n \times m$ dictionary $\mathbf{A}$ and $m$-dimensional codes $\mathbf{x}_i$ contained (up to noise) in the appropriately scaled $n \times \hat m$ dictionary $\mathbf{B}$ and $\hat m$-dimensional codes $\mathbf{\hat x}_i$. In particular, when $\hat m = m$ all columns of $\mathbf{A}$ and coefficients of each $\mathbf{x}_i$ recoverable, as per Def.~\ref{maindef}. For expositional clarity, we delay defining the (explicit) constant $C_1 = C_1(\mathbf{A}, \{\mathbf{x}_i\}_{i=1}^N, E)$ until Section \ref{DUT} (\eqref{Cdef}), since it requires additional definitions not critical to the statement of our results.

@@

\begin{corollary}\label{DeterministicUniquenessCorollary}
Given $n, m$, $k < m$, and a regular hypergraph $E \subseteq 2^{[m]}$ of rank $k$ satisfying the singleton intersection property, there are $N =  |E| \left[ (k-1){m \choose k}+1 \right]$ vectors \mbox{$\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$} such that every matrix $\mathbf{A} \in \mathbb{R}^{n \times m}$ satisfying \eqref{SparkCondition} generates a dataset $Y = \{\mathbf{A}\mathbf{x}_1, \ldots, \mathbf{A}\mathbf{x}_N\}$ with a stable $k$-sparse representation in $m$ dimensions.
\end{corollary}
\begin{proof}
Something quick here.
\end{proof}

@@

It is straightforward to provide a probabilistic extension of Thm.~\ref{DeterministicUniquenessTheorem} using the following fact in random matrix theory.  A random matrix $\mathbf{A} \in \mathbb{R}^{n \times m}$ satisfies \eqref{SparkCondition} with probability one (or with ``high probability'' for discrete variables) 
provided:
\begin{align}\label{CScondition}
n \geq \gamma k\log\left(\frac{m}{k}\right),
\end{align}
%
where $\gamma$ is a positive constant that depends on the particular distribution from which the entries of $\mathbf{A}$ are sampled i.i.d. (many ensembles suffice, e.g. \cite[Sec.~4]{Baraniuk08}).

In fact, the spark condition can be made explicit.  Let $\mathbf{A}$  be the $n \times m$ matrix of $nm$ indeterminates $A_{ij}$. When real numbers are substituted for $A_{ij}$, the resulting matrix satisfies \eqref{SparkCondition} if and only if the following polynomial is nonzero:
\begin{align*}
f(\mathbf{A}) := \prod_{S \in {[m] \choose k}} \sum_{S' \in {[n] \choose k}} (\det \mathbf{A}_{S',S})^2,
\end{align*}
%
where for any $S' \in {[n] \choose k}$ and $S \in {[m] \choose k}$, the symbol $\mathbf{A}_{S',S}$ denotes the submatrix of entries $A_{ij}$ with $(i,j) \in S' \times S$.   We note that the large number of terms in this product is likely necessary due to the NP-hardness of deciding whether a given matrix $A$ satisfies the spark condition \cite{tillmann2014computational}.

Since $f$ is a real analytic function, one substitution of real numbers with $f(\mathbf{A}) \neq 0$ implies that its zeroes form a set with measure zero. Hence, almost every $n \times m$ real matrix $\mathbf{A}$ satisfies \eqref{SparkCondition} provided \eqref{CScondition} holds for a value of $\gamma$ for some distribution. We remark that the precise relationship between $m$, $n$, and $k$ guaranteeing that $f$ is not identically zero is a challenging problem in real algebraic geometry. In any case, we set $\gamma_0$ to be the smallest known such $\gamma$.

>>>>

% an important contribution of our work is to
Next, we identify the combinatorial criteria on support sets of generating codes allowing for stable sparse 
representations.  Let $[m]$ denote the set $\{1, \ldots, m\}$, its power set $2^{[m]}$, and ${[m] \choose k}$ the set of subsets of $[m]$ of size $k$. Let $\text{\rm Span}\{\mathbf{v}_1, \ldots, \mathbf{v}_\ell\}$ be the $\mathbb{R}$-linear span of vectors $\mathbf{v}_1, \ldots, \mathbf{v}_\ell$. A vector $\mathbf{x}$ is said to be \emph{supported} on $S \subseteq [m]$, written supp$(\mathbf{x}) = S$, when $\mathbf{x} \in \text{\rm Span} \{\mathbf{e}_j\}_{j\in S}$, where $\mathbf{e}_j$ are the standard basis elements in $\mathbb R^m$.  Let $\mathbf{M}_j$ be the $j$th column of a matrix $\mathbf{M}$ and similarly let $\mathbf{M}_S$ be the submatrix formed by the columns of $\mathbf{M}$ indexed by $S$ (we set $\text{\rm Span}\{\mathbf{M}_\emptyset\} := \{\textbf{0}\}$).  We say a hypergraph $E \subseteq 2^{[m]}$ has \textit{rank} $k$ when every $S \in E$ has cardinality $|S| \leq k$ (and \textit{$k$-uniform} when equality holds). We also say $E$ is \emph{regular} when every element of $[m]$ is contained in exactly $\ell$ elements of $E$ for some $\ell > 0$ (for given $\ell$, we say $E$ is $\ell$-regular).

\begin{definition}
Given a hypergraph $E \subseteq 2^{[m]}$, the \textbf{star} $F(i)$ centered at $i$ is the collection of $S \in E$ containing $i$. We say $E$ satisfies the \textbf{singleton intersection property} (\textbf{SIP}) when $\cap_{S \in F(i)} S = \{i\}$ for all $i \in [m]$. 
\end{definition}

@@

\begin{theorem}\label{DeterministicUniquenessTheorem}
% Fix positive integers $k, \ell, m, m', n, p$ satisfying $k < m$ and $p \leq m \leq m' \leq (\ell m - p) / (\ell - 1)$. 
Fix integers $n, \ell$, and $k < m \leq \hat m$. Suppose $\mathbf{A}~\in~\mathbb{R}^{n \times m}$ satisfies the spark condition \eqref{SparkCondition} and that $k$-sparse \mbox{$\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$} contain at least \mbox{$(k-1){\hat m \choose k}+1$} vectors supported in general linear position on each set in some $\ell$-regular $E \subseteq {[m] \choose k}$ satisfying the SIP.  There exists a constant $C_1 > 0$ for which the following holds for all $\varepsilon < L_2(\mathbf{A}) / C_1$.

@@

In other words, the smaller the difference $\hat m - m$, the more columns and coefficients of the original $n \times m$ dictionary $\mathbf{A}$ and $m$-dimensional codes $\mathbf{x}_i$ contained (up to noise) in the appropriately scaled $n \times \hat m$ dictionary $\mathbf{B}$ and $\hat m$-dimensional codes $\mathbf{\hat x}_i$. In particular, when $\hat m = m$ all columns of $\mathbf{A}$ and coefficients of each $\mathbf{x}_i$ recoverable, as per Def.~\ref{maindef}. For expositional clarity, we delay defining the (explicit) constant $C_1 = C_1(\mathbf{A}, \{\mathbf{x}_i\}_{i=1}^N, E)$ until Section \ref{DUT} (\eqref{Cdef}).  % , since it requires additional definitions not critical to the statement of our results.

@@

\begin{corollary}\label{DeterministicUniquenessCorollary}
Given $n, m$, $k < m$, and a regular hypergraph $E \subseteq {[m] \choose k}$ of rank $k$ with the SIP, there are $N =  |E| \left[ (k-1){m \choose k}+1 \right]$ vectors \mbox{$\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$} such that every matrix $\mathbf{A} \in \mathbb{R}^{n \times m}$ satisfying \eqref{SparkCondition} generates a dataset $Y = \{\mathbf{A}\mathbf{x}_1, \ldots, \mathbf{A}\mathbf{x}_N\}$ with a stable $k$-sparse representation in $m$ dimensions.
\end{corollary}
\begin{proof}
Something quick here.
\end{proof}

@@

It is straightforward to provide a probabilistic extension of Thm.~\ref{DeterministicUniquenessTheorem} using the following fact in random matrix theory.  A random matrix $\mathbf{A} \in \mathbb{R}^{n \times m}$ satisfies \eqref{SparkCondition} with probability one (or with ``high probability'' for discrete variables) 
provided:
\begin{align}\label{CScondition}
n \geq \gamma k\log\left(\frac{m}{k}\right),
\end{align}
%
in which $\gamma$ is a positive constant that depends on the particular distribution from which the entries of $\mathbf{A}$ are sampled i.i.d. (many ensembles suffice, e.g. \cite[Sec.~4]{Baraniuk08}).

We motivate our next theorem by the following observation:  the spark condition can be made explicit.  Let $\mathbf{A}$  be the $n \times m$ matrix of $nm$ indeterminates $A_{ij}$. When real numbers are substituted for $A_{ij}$, the resulting matrix satisfies \eqref{SparkCondition} if and only if the following polynomial is nonzero:
\begin{align*}
f(\mathbf{A}) := \prod_{S \in {[m] \choose k}} \sum_{S' \in {[n] \choose k}} (\det \mathbf{A}_{S',S})^2,
\end{align*}
%
where for any $S' \in {[n] \choose k}$ and $S \in {[m] \choose k}$, the symbol $\mathbf{A}_{S',S}$ denotes the submatrix of entries $A_{ij}$ with $(i,j) \in S' \times S$.   We note that the large number of terms in this product is likely necessary due to the NP-hardness of deciding whether a given matrix $A$ satisfies the spark condition \cite{tillmann2014computational}.

Since $f$ is a real analytic function, one substitution of real numbers with $f(\mathbf{A}) \neq 0$ implies that its zeroes form a set of (Borel) measure zero. Hence, almost every $n \times m$ real matrix $\mathbf{A}$ satisfies \eqref{SparkCondition} provided \eqref{CScondition} holds for a value of $\gamma$ for some distribution. We remark that the precise relationship between $m$, $n$, and $k$ guaranteeing that $f$ is not identically zero is a challenging problem in real algebraic geometry. In any case, we set $\gamma_0$ to be the smallest known such $\gamma$.
